\section{The Bytes Memory Model}
\label{sec:bytes-model}

The behavior of the Bytes memory model is essentially the same as the Typed
memory model except that, while Typed dispatches the aliased memory according
to the types of the values, the Bytes memory model has a single memory which
is an array from pointer base to memory block, memory blocks being arrays from
pointer offset to byte.

\paragraph{Modeling.} When a value is written to the memory, it first encoded
as a sequence of bytes. More precisely, unsigned integers are directly encoded
into bytes via modulo and division. For other fundamental types, WP first
translate them into an unsigned and then encode the obtained unsigned, the
following applies for the translation to unsigned:
\begin{itemize}
  \item signed integers are translated using (morally) an ACSL cast,
  \item for pointers, we use the physical translation axiomatized in WP,
  \item for float, we assume that there exists such a function without providing it.
\end{itemize}
The sequence of bytes that we obtained is then written bytes by bytes in memory.

When a value is read from the memory, the opposite applies: we read the bytes
from the memory, build the unsigned value and then translate it back to the
expected type.

This modeling is \textbf{super costly} for proof, since the read and writes
involves the encoding + decoding of values. It makes formulas quite complex.

\paragraph{Mitigation.} There are two main mitigations.

First, we provide a lot of lemmas about load/store operations so that when the
code does not exhibit heterogeneous load/store, the SMT solvers do not have to
reason about the encoding. Furthermore, there are also shortcuts for load/store
when the only heterogeneous behavior is the presence of a sign or not.

Second, we use the Hoare memory model when there is no alias. It avoids modeling
bytes for every single variable of the program.

\paragraph{Limitations.} The model has three main limitations currently:
\begin{itemize}
  \item it assumes little-endian,
  \item it cannot really deal with low-level operations for pointers and floats,
        because WP assumes that a conversion function exists (which makes sense)
        but does not exhibit this function,
  \item in default mode (+ Hoare), if a union is not aliased it is \textbf{not}
        sent into the Bytes model but into the Hoare one that does not take in
        account heterogeneous operations. One can use \texttt{+raw} to use the
        Bytes memory model everywhere, but be prepared to the fact that
        properties will be extremely hard to prove.
\end{itemize}
